---
title: "get_data_from_gbif"
author: "Sander Devisscher"
date: "`r Sys.Date()`"
output: html_document
---

```{r libraries}
library(rgbif)
library(dplyr)
library(sf)
library(readr)
library(magrittr)
```

get the polygon of flanders and convert it into a wkt, this will be used to
limit the rgbif query to flanders.

```{r get flanders geojson}
sf_use_s2(TRUE)

bel <- st_read("./data/output/UAT_processing/grid/gewestbel.shp") %>% 
  st_simplify(dTolerance = 10) %>% 
  st_make_valid() %>% 
  st_union() %>% 
  st_as_sf() %>% 
  st_simplify(dTolerance = 1500) %>%
  st_cast(to = "POLYGON") %>% 
  st_as_sfc() %>% 
  st_as_text() 


bel <- gsub(pattern = ", NaN", 
            replacement = "",
            bel)
```

The chunk below is not necessary for the script to run, 
but it is useful to check if the polygon is correct.
To run the chunk below you should **not** run `st_as_sfc() %>% st_as_text()` 
while creating the `bel` object in the chunk above.


```{r test bel, eval=FALSE}
library(leaflet)

leaflet(bel) %>% 
  addTiles() %>% 
  addPolylines()
```


Waarnemingen.be - Non-native animal occurrences in Flanders and the Brussels Capital Region, Belgium: https://www.gbif.org/dataset/9a0b66df-7535-4f28-9f4e-5bc11b8b096c

iNaturalist Research-grade Observations: https://www.gbif.org/dataset/50c9509d-22c7-4a22-a47d-8c48425ef4a7

```{r download data from gbif}
taxonkeys <- c(1311477)

gbif_downloadKey <- occ_download(pred_in("taxonKey", taxonkeys),
                                 pred_within(bel),
                                 user = Sys.getenv("gbif_user"),
                                 pwd = Sys.getenv("gbif_pwd"),
                                 email = Sys.getenv("email"),
                                 curlopts = list(verbose = TRUE,
                                                 http_version = 2,
                                                 forbid_reuse = TRUE))

occ_download_wait(gbif_downloadKey, 
                  curlopts = list(verbose = TRUE,
                                  http_version = 2,
                                  forbid_reuse = TRUE))

gbif_download <- occ_download_get(gbif_downloadKey,
                                  path = tempdir(),
                                  overwrite = TRUE,
                                  curlopts = list(verbose = TRUE,
                                                  http_version = 2,
                                                  forbid_reuse = TRUE))

rawdata <- occ_download_import(x = gbif_download) %>% 
  write_csv2("./data/input/Vespa-Watch_gbif.csv")
```
