---
title: "processing translation files"
author: "Sander Devisscher"
date: "2025-09-04"
output: html_document
---

This script is tiggered by changes in *translations*, *translations_simple* 
and/or *translations_regions*.
Currently a transition period these files reside in *./data/output/uat_direct/*.
In the future these files will be moved to *./data/interim/translations/*

This script converts the files into 1 files per language and stores the results 
as *translations_nl*, *translations_fr* & *translations_en* in 
*./data/output/uat_direct/translations*.

```{r libraries}
library(readr)
library(dplyr)
```

```{r determine input directory}
new_dir <- "./data/interim/translations/"
old_dir <- "./data/output/uat_direct/"

if(all(file.exists(paste0(new_dir, "translations.csv")),
       file.exists(paste0(new_dir, "translations_simple.csv")),
       file.exists(paste0(new_dir, "translations_regions.csv")))){
  current_dir <- new_dir
}else{
  current_dir <- old_dir
}

cat("reading data from:", current_dir, "\n")
```

```{r set output dir}
output_dir <- "./data/output/uat_direct/translations/"
```

```{r read files}
translations_raw <- read_csv2(paste0(current_dir, "translations.csv"))

translations_simple_raw <- read_csv2(paste0(current_dir, "translations_simple.csv"))

translations_regions_raw <- read_csv2(paste0(current_dir, "translations_regions.csv"))

translations_template <- read_csv(paste0(output_dir, "translation_nl.csv"), n_max = 5)
```

