---
title: "processing translation files"
author: "Sander Devisscher"
date: "2025-09-04"
output: html_document
---

This script is triggered by changes in *translations*, *translations_simple* 
and/or *translations_regions*.
Currently these files reside in *./data/output/uat_direct/*.
In the future these files will be moved to *./data/interim/translations/*

This script converts the files into 1 file per language and stores the results 
as *translations_nl*, *translations_fr* & *translations_en* in 
*./data/output/uat_direct/translations*.

```{r libraries}
library(readr)
library(dplyr)
library(tidyr)
```

```{r determine input directory}
new_dir <- "./data/interim/translations/"
old_dir <- "./data/output/UAT_direct/"

if(all(file.exists(paste0(new_dir, "translations.csv")),
       file.exists(paste0(new_dir, "translations_simple.csv")),
       file.exists(paste0(new_dir, "translations_regions.csv")))){
  current_dir <- new_dir
}else{
  current_dir <- old_dir
}

cat("reading data from:", current_dir, "\n")
```

```{r set output dir}
output_dir <- "./data/output/UAT_direct/translations/"
```

```{r read template, eval=FALSE}
translations_template <- read_csv(paste0(output_dir, "translation_nl.csv"))
```

```{r convert files & combine them}
files <- c("translations", "translations_simple", "translations_regions")

translations_combined <- data.frame()

for(f in files){
  temp_raw <- read_csv2(paste0(current_dir, f, ".csv"))
  
  temp <- temp_raw %>% 
    pivot_longer(cols = -title_id,
                 names_to = "type_language",
                 values_to = "value") %>% 
    separate(col = type_language,
             into = c("type", "language")) %>% 
    mutate(id = paste(title_id, type, sep = "_")) %>% 
    select(id, value, language)
  
  if(nrow(translations_combined) == 0){
    translations_combined <- temp
  }else{
    translations_combined <- rbind(translations_combined, temp)
  }
}

translations_combined <- translations_combined %>% 
  distinct() 
```

```{r split into languages}
languages <- c("en", "nl", "fr")

for(l in languages){
  translations_language <- translations_combined %>% 
    filter(language == l) %>% 
    arrange(id)
  
  translations_language[[l]] <- translations_language$value
  
  translations_language <- translations_language %>% 
    select(id, all_of(l)) %>% 
    write_csv(paste0(output_dir, "translation_", l, ".csv"))
}
```
