---
title: "create_new_links"
author: "Sander Devisscher"
date: "2025-05-12"
output: html_document
---

```{r libraries}
library(RCurl)
library(tidyverse)
```

```{r functies}
source("./HTML_pages/Render_single_HTML.R")
source("./src/assign_dutch_article.R")
```

```{r read template}
links_template <- paste(readLines("./data/input/links_template.txt"), collapse = "\n")
```

```{r read GRIIS checklist}
griis_checklist_raw <- read_tsv("./data/output/UAT_processing/data_input_checklist_indicators.tsv")
```

```{r read faunabeheer species}
#Test if the faunabeheer species file exists in the reporting-rshiny-grofwildjacht 
# repository on the 557-make-species-page-linkable-by-vbpexotenportaal-inward-linking 
# branch, if not, read the file from the uat branch.

tryCatch({
  faunabeheer_species <- read_csv("https://raw.githubusercontent.com/inbo/reporting-rshiny-grofwildjacht/557-make-species-page-linkable-by-vbpexotenportaal-inward-linking/reporting-grofwild/inst/extdata/species-info.csv")
}, error = function(e) {
  message("File not found on the 557-make-species-page-linkable-by-vbpexotenportaal-inward-linking branch, reading from uat branch.")
  faunabeheer_species <- read_csv("https://raw.githubusercontent.com/inbo/reporting-rshiny-grofwildjacht/uat/reporting-grofwild/inst/extdata/species-info.csv")
})

if(!"gbifkey" %in% colnames(faunabeheer_species)){
  stop("The faunabeheer species file does not contain a 'gbifkey' column. Please check the file.")
}

faunabeheer_species <- faunabeheer_species %>% 
  mutate(faunabeheer_link1 = paste0("https://faunabeheer.inbo.be/app/01_faunabeheer/#", gsub(" ", "%20", `species name`)),
         faunabeheer_link2 = paste0("https://faunabeheer.inbo.be/app/01_faunabeheer/#", gbifkey))

# test both urls and write the first that works into faunabeheer_link
# This code anticipates changes in the faunabeheer links, where the first link is the one with the species name and the second link is the one with the gbifkey.
faunabeheer_species <- faunabeheer_species %>% 
  mutate(faunabeheer_link = case_when(
    RCurl::url.exists(faunabeheer_link1) ~ faunabeheer_link1,
    RCurl::url.exists(faunabeheer_link2) ~ faunabeheer_link2,
    TRUE ~ NA_character_)) %>% 
  select(gbifkey, faunabeheer_link)
```

```{r detect images}
pngs <- list.files(path = "./HTML_pages/Pictures/", 
                   pattern = ".png", 
                   full.names = FALSE)
jpgs <- list.files(path = "./HTML_pages/Pictures/", 
                   pattern = ".jpg", 
                   full.names = FALSE)
jpegs <- list.files(path = "./HTML_pages/Pictures/", 
                    pattern = ".jpeg", 
                    full.names = FALSE)

image_files <- c(pngs, jpgs, jpegs)

image_files <- image_files %>% 
  as.data.frame() %>% 
  rename(image_file = ".") %>% 
  separate(image_file, sep = "_", into = c("taxonKey", "copyright"), extra = "drop") %>% 
  separate(copyright, sep = "\\.", into = c("copyright", "ext"), extra = "drop") %>%
  mutate(taxonKey = as.numeric(taxonKey)) %>%
  filter(!is.na(taxonKey))
```

```{r determine custom links pages}
custom_links_pages <- dir(path = "./HTML_pages/Rmd_files/", pattern = "[0-9].Rmd") %>% 
  gsub(pattern = ".Rmd", replacement = "") %>% 
  as.integer()
```

Most species have "het" as the correct article in dutch while occasionally 
"de" is used, these should be set manually by adding the nubkey of these species
to the "de_species" - list. 

```{r assign dutch article}
het_exception_list <- readLines("./data/interim/links_html_het_list.txt")
de_exception_list <- readLines("./data/interim/links_html_de_list.txt")

griis_checklist_raw <- assign_dutch_article(griis_checklist_raw,
                                            het_exception = c(5403508, 8288930, 5220136, het_exception_list),
                                            de_exception = c(2874199, de_exception_list))
```

```{r prep griis checklist}
griis_checklist <- griis_checklist_raw %>% 
  filter(!is.na(nubKey) & !nubKey %in% custom_links_pages) %>% 
  select(nubKey, scientificName = canonicalName, vernacular_name_nl, vernacular_name_en, vernacular_name_fr, article_nl) %>% 
  distinct() %>% 
  separate(vernacular_name_nl, sep = ", ", into = "vernacular_name_nl", extra = "drop") %>% 
  separate(vernacular_name_en, sep = ", ", into = "vernacular_name_en", extra = "drop") %>%
  separate(vernacular_name_fr, sep = ", ", into = "vernacular_name_fr", extra = "drop") %>% 
  mutate(vernacular_name_nl2 = case_when(!is.na(vernacular_name_nl) ~ paste(article_nl, tolower(vernacular_name_nl)),
                                         TRUE ~ vernacular_name_nl),
         vernacular_name_en2  = case_when(!is.na(vernacular_name_en) ~ paste("the", vernacular_name_en),
                                         TRUE ~ vernacular_name_en),
         vernacular_name_fr2 = tolower(vernacular_name_fr)) %>% 
  mutate(vernacular_name_nl = coalesce(vernacular_name_nl, paste0("<i>", scientificName, "</i>")),
         vernacular_name_en = coalesce(vernacular_name_en, paste0("<i>", scientificName, "</i>")),
         vernacular_name_fr = coalesce(vernacular_name_fr, paste0("<i>", scientificName, "</i>")),
         vernacular_name_nl2 = coalesce(vernacular_name_nl2, paste0("<i>", scientificName, "</i>")),
         vernacular_name_en2 = coalesce(vernacular_name_en2, paste0("<i>", scientificName, "</i>")),
         vernacular_name_fr2 = coalesce(vernacular_name_fr2, paste0("<i>", scientificName, "</i>"))) %>% 
  left_join(faunabeheer_species, by = c("nubKey" = "gbifkey")) %>%
  mutate(faunabeheer_link_nl = case_when(nubKey %in% faunabeheer_species$gbifkey ~ 
                                           paste0('<--VERNACULAR_NAME_NL2--> is in Vlaanderen geclassificeerd als jachtwildsoort, meer info met betrekking tot het beheer en schade in Vlaanderen vindt u terug op de <a href=\\\\"', faunabeheer_link, '\\\\"target=\\\\"_blank\\\\">faunabeheer pagina</a>.'),
                                         TRUE ~ ""),
         faunabeheer_link_en = case_when(nubKey %in% faunabeheer_species$gbifkey ~ 
                                           paste0('<--VERNACULAR_NAME_EN--> is classified as a game species in Flanders, more information concerning management and damages in Flanders can be found on the <a href=\\\\"', faunabeheer_link, '\\\\"target=\\\\"_blank\\\\">wildlife management page</a> (note that this information is currently only available in Dutch)'),
                                         TRUE ~ ""),
         faunabeheer_link_fr = case_when(nubKey %in% faunabeheer_species$gbifkey ~ 
                                           paste0('<--VERNACULAR_NAME_FR--> est classé comme espèce gibier en Flandre. Vous trouverez plus d&apos;informations sur la gestion et les dommages en Flandre sur la page <a href=\\\\"', faunabeheer_link, '\\\\"target=\\\\"_blank\\\\">gestion de la faune sauvage</a> (à noter que ces informations ne sont actuellement disponibles qu` en néerlandais)'),
                                         TRUE ~ "")) %>% 
  left_join(image_files, by = c("nubKey" = "taxonKey")) %>% 
  mutate(image_copyright = case_when(!is.na(copyright) ~ 
                                       paste0(nubKey, "_", copyright, ".", ext),
                                     TRUE ~ ""),
         image_link = case_when(!is.na(copyright) ~ paste0('<img src=\\\\"", image_link, "\\\\" width=\\\\"", width, "\\\\" height=\\\\"", height, "\\\\"><br>
<span style=\\\\"", copyright_font, "\\\\">© <--COPYRIGHT--></span><br>'),
TRUE ~ "")) %>% 
  distinct()
```

```{r Convert geographical annotations to Uppercase}
geo_words = c("Chinese", 
              "Amerikaanse", 
              "Europees", 
              "Europese", 
              "Japans", 
              "Afrikaanse", 
              "Indische", 
              "Canadese", 
              "Amerikaans", 
              "Australische",
              "Aziatische",
              "Siberische", 
              "Oosterse", 
              "Westelijke", 
              "Noordse",
              "Noorse",
              "Zuidelijke", 
              "Californische", 
              "Carolina",
              "Mississippi",
              "Griekse", 
              "Magelhaengans")

for(g_up in geo_words){
  g_low <- tolower(g_up)
  
  cat("Converting", g_low, "into", g_up, "\n")
  
  griis_checklist <- griis_checklist %>% 
    mutate(vernacular_name_nl2 = str_replace_all(vernacular_name_nl2, 
                                                 pattern = g_low, 
                                                 replacement = g_up))
}
```

```{r redo all?}
redo_all <- FALSE # Set to TRUE to redo all files, or FALSE to skip existing files
redo_existing <- FALSE # Set to TRUE to redo only existing files, or FALSE to redo all
```

| ALL = T & EXI. = F 	| All on GRIIS checklist (inclu. non - existing but exclu. Custom) 	|
| ALL = T & EXI. = T 	| All existing on GRIIS checklist (exclu. Custom) 	|
| ALL = F & EXI. = T 	| Skip all => no changes 	|
| ALL = F & EXI. = F 	| Only non - existing files (exclu. Custom) 	|

```{r determine existing}
if(redo_existing){
  existing_files <- dir(path = "./HTML_pages/Rmd_files/generic_links_pages/", 
                        pattern = "[0-9].Rmd") %>% 
    gsub(pattern = ".Rmd", replacement = "") %>% 
    as.integer()
  
  griis_checklist <- griis_checklist %>% 
    filter(nubKey %in% existing_files)
}
```

```{r sample GRIIS checklist for troubleshooting, eval=FALSE}
griis_checklist <- sample_n(griis_checklist, 5, replace = TRUE)
```

```{r create link files}
for(n in 1:nrow(griis_checklist)){
  
  # get the current row
  current_row <- griis_checklist[n, ]
  
  # create filename
  filename <- paste0("./HTML_pages/Rmd_files/generic_links_pages/", current_row$nubKey, ".Rmd")
  
  # check if the file for the current row already exists
  if(file.exists(filename) & !redo_all){
    
    print(paste0("File already exists for ", current_row$scientificName, ", skipping..."))
    
    next
  }
  
  # replace the placeholders in the template with actual values
  new_link <- links_template %>% 
    gsub(pattern = "<--TAXONKEY-->", 
         replacement = as.character(current_row$nubKey), 
         x = .) %>% 
    gsub(pattern = "<--SCIENTIFICNAME-->",
         replacement = current_row$scientificName, 
         x = .) %>% 
    gsub(pattern = "<--FAUNABEHEER_LINK_NL-->",
         replacement = current_row$faunabeheer_link_nl, 
         x = .) %>%
    gsub(pattern = "<--FAUNABEHEER_LINK_EN-->",
         replacement = current_row$faunabeheer_link_en, 
         x = .) %>%
    gsub(pattern = "<--FAUNABEHEER_LINK_FR-->",
         replacement = current_row$faunabeheer_link_fr, 
         x = .) %>%
    gsub(pattern = "<--VERNACULAR_NAME_NL-->",
         replacement = current_row$vernacular_name_nl, 
         x = .) %>%
    gsub(pattern = "<--VERNACULAR_NAME_NL2-->",
         replacement = current_row$vernacular_name_nl2, 
         x = .) %>%
    gsub(pattern = "<--VERNACULAR_NAME_EN-->",
         replacement = current_row$vernacular_name_en, 
         x = .) %>%
    gsub(pattern = "<--VERNACULAR_NAME_EN2-->",
         replacement = current_row$vernacular_name_en2, 
         x = .) %>%
    gsub(pattern = "<--VERNACULAR_NAME_FR-->",
         replacement = current_row$vernacular_name_fr, 
         x = .) %>%
    gsub(pattern = "<--VERNACULAR_NAME_FR2-->",
         replacement = current_row$vernacular_name_fr2, 
         x = .) %>%
    gsub(pattern = "<--IMAGE-COPYRIGHT-->",
         replacement = current_row$image_copyright, 
         x = .) %>%
    gsub(pattern = "<--IMAGE_LINK-->",
         replacement = current_row$image_link, 
         x = .) %>% 
    gsub(pattern = "<--COPYRIGHT-->",
         replacement = current_row$copyright, 
         x = .)
  
  
  # write the new link to a file
  writeLines(new_link, filename)
  
  render_single_html(input_file = filename,
                     output_path = "../../HTML/")
}
```

