---
title: "Cube pre-processing"
author:
- Damiano Oldoni
- Toon Van Daele
- Tim Adriaens
- Sander Devisscher
- Soria Delva
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This script is based on [05_occurrence_indicators_preprocessing.Rmd](https://github.com/trias-project/indicators/blob/main/src/05_occurrence_indicators_preprocessing.Rmd) from the [trias-project/indicators](https://github.com/trias-project/indicators) repository.
It has been adapted to the specific needs of the alien species portal.
This document describes the pre-processing of occurrence data of alien species. 
The outputs of this pipeline will be used as input for modelling.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

Load libraries:

```{r load_libraries, message = FALSE}
library(dplyr) # To do data science
library(readr)
library(magrittr)
library(purrr)
library(stringr)
library(tidyr)
library(tidylog) # To provide feedback on dplyr functions
library(progress) # To add progress bars
library(here) # To find files
library(lubridate) # To work with dates
library(rgbif) # To get taxa from publisehd unified checklist
library(sf) # To work with spatial data
library(testthat) # To test functions
```

```{r load correct version of INBOtheme, eval=TRUE}
if(Sys.getenv("amiabot") != "yes"){
  print("Executor of the script is a human >> checking INBOtheme version")
  # determine the loaded INBOtheme version
  loaded_inbotheme <- installed.packages() %>% 
    as_tibble() %>% 
    filter(Package == "INBOtheme")
  loaded_inbotheme_version <- loaded_inbotheme$Version
  
  if(is_empty(loaded_inbotheme_version)){
    print("INBOtheme is not installed >> installing INBOtheme")
    devtools::install_github("inbo/INBOtheme@v0.5.8")
  }
  
  if(length(loaded_inbotheme_version) > 1 || loaded_inbotheme_version != "0.5.8"){
    print("INBOtheme version is not correct >> loading correct version")
    # remove installed INBOtheme
    remove.packages("INBOtheme")
    
    # load correct version of INBOtheme
    devtools::install_github("inbo/INBOtheme@v0.5.8")
  }else{
    print("Correct INBOtheme version is loaded")
  }
}else{
  print("Executor of the script is a bot >> correct INBOtheme version is loaded")
}

library(INBOtheme)
```

# Connect to bucket
## set credentials
```{r set creds}
Sys.setenv("AWS_DEFAULT_REGION" = "eu-west-1")

if(Sys.getenv("S3_BUCKET") != ""){
  UAT_bucket <- Sys.getenv("S3_BUCKET")
}else{
  UAT_bucket <- Sys.getenv("UAT_bucket")
}
```

## connect to bucket
```{r connect to bucket}
if(Sys.getenv("amiabot") != "yes"){
  print("Executor of the script is a human >> connecting to bucket")
  source("./src/connect_to_bucket.R")
  connect_to_bucket(bucket_name = UAT_bucket)
}
```

# Get data

## read utm1_bel_with_regions file
```{r read utm1_gemeentes_provincies}
utm1_gemeentes_provincies <- sf::st_read("./data/output/UAT_processing/grid/utm1_bel_with_regions.gpkg", 
                                         layer="utm1_bel_with_regions" )
```

We get occurrence data as obtained in `trias-project/occ-processing` repository. 

## Occurrence data

### Alien species

We load occurrence data related to alien species in Belgium:

```{r read_data_cube}
df <- read_csv("./cube_data/be_alientaxa_cube.csv") 

print(file.info("./cube_data/be_alientaxa_cube.csv"))
```

To save the data cube to the S3 bucket, we use the following code:
A temporary directory is created to store the data cube.
```{r create_temp_dir}
cube_path <- tempdir()
dir.create(cube_path, showWarnings = FALSE)
```

For display on invasion history maps of the alienspecies portal we add the 
gemeente and provincie to the data cube.
This is done by joining the data cube with the `utm1_gemeentes_provincies` dataset.
which is updated by the following code:
[processing_utm10_files.R](./src/Processing/processing_utm10_files.R)

```{r add gemeente and provincie to data_cube}
be_alientaxa_cube <- df %>% 
  left_join(utm1_gemeentes_provincies, by = c("eea_cell_code" = "CELLCODE")) %>% 
  st_drop_geometry() %>% 
  filter(!is.na(isFlanders)|!is.na(isWallonia)|!is.na(isBrussels))

be_alientaxa_cube %>%
  select(c(year,eea_cell_code,taxonKey,n = obs,min_coord_uncertainty,isFlanders,isWallonia,isBrussels,gemeente,provincie,gewest)) %>% 
  write_csv(file.path(cube_path, "be_alientaxa_cube.csv"))
```

Recalculate last_observed column per taxonkey
```{r recalculate_last_observed}
last_observed_taxon <- be_alientaxa_cube %>% 
  group_by(taxonKey) %>% 
  summarise(last_observed_recalc = max(year, na.rm = TRUE),
            isFlanders = any(isFlanders),
            isWallonia = any(isWallonia),
            isBrussels = any(isBrussels),
            isBelgium = TRUE) %>% 
  ungroup() %>% 
  pivot_longer(cols = c(isFlanders, isWallonia, isBrussels, isBelgium), names_to = "region", values_to = "isRegion") %>% 
  filter(isRegion) %>% 
  select(-isRegion) %>% 
  mutate(region = case_match(region,
                             "isFlanders" ~ "Flemish Region",
                             "isWallonia" ~ "Walloon Region",
                             "isBrussels" ~ "Brussels-Capital Region",
                             "isBelgium" ~ "Belgium")) 

data_input_checklist_indicators <- read_tsv("./data/output/UAT_processing/data_input_checklist_indicators.tsv") %>% 
  left_join(last_observed_taxon, by = c("nubKey" = "taxonKey", "locality" = "region")) %>% 
  mutate(last_observed = case_when(is.na(last_observed) ~ last_observed_recalc,
                                   last_observed < last_observed_recalc ~ last_observed_recalc,
                                   TRUE ~ last_observed)) %>% 
  select(-last_observed_recalc) %>% 
  write_tsv("./data/output/UAT_processing/data_input_checklist_indicators.tsv")
```

Free some ram. `be_alientaxa_cube` is no longer needed.
```{r free_ram}
df <- be_alientaxa_cube

rm(be_alientaxa_cube)
gc() #garbage collection
```

The function `createTabularData()` from the alien species portal package is used 
to process & save the data cube to the S3 bucket for use in the alien species portal.

```{r get pre update values}
cube_pre_update <- as.data.frame(aws.s3::get_bucket(bucket = UAT_bucket)) %>% 
  filter(grepl("be_alientaxa_cube_processed.parquet", Key))

cube_pre_update_lastmod <- cube_pre_update$LastModified
cube_pre_update_size <- cube_pre_update$Size
```

```{r save data_cube to S3 bucket}
alienSpecies::createTabularData(dataDir = cube_path, type = "occurrence", bucket = UAT_bucket) 
```

```{r get post update values}
cube_post_update <- as.data.frame(aws.s3::get_bucket(bucket = UAT_bucket)) %>% 
  filter(grepl("be_alientaxa_cube_processed.parquet", Key))

cube_post_update_lastmod <- cube_post_update$LastModified
cube_post_update_size <- cube_post_update$Size

cube_post_update_nrow <- nrow(alienSpecies::loadTabularData(type = "occurrence"))
```

We test if the data have been correctly saved:
```{r test_load_occurrenceData}
test_that("Load occurrenceData", { 
  occurrenceData <- alienSpecies::loadTabularData(type = "occurrence") 
  expect_s3_class( occurrenceData, "data.table") 
}) 

print("lastmod pre update")
print(cube_pre_update_lastmod)
print("lastmod post update")
print(cube_post_update_lastmod)

print("size pre update")
print(cube_pre_update_size)
print("size post update")
print(cube_post_update_size)

print("nrow update")
print(nrow(df %>% filter(year >= 1950 & !is.na(year))))
print("nrow post update")
print(cube_post_update_nrow)

# we expect the post update last modified date to be later than the pre update last modified date
test_that("UPLOAD FAILED: last modified date not updated",
          {
            expect_true(cube_post_update_lastmod > cube_pre_update_lastmod)
          })

# we expect the post update number of rows to be equal to the new number of rows
test_that("UPLOAD FAILED: number of rows not equal",
          {
            expect_equal(cube_post_update_nrow, nrow(df %>% filter(year >= 1950 & !is.na(year))))
          })

df_species <- df %>% 
  filter(year >= 1950 & !is.na(year)) %>% 
  distinct(taxonKey) %>% 
  pull()

post_update_species <- alienSpecies::loadTabularData(type = "occurrence") %>%
  distinct(taxonKey) %>%
  pull()

# we expect all species included in dataset provided to be included in the uploaded data
test_that("UPLOAD FAILED: not all species included",
          {
            
            
            expect_in(df_species, post_update_species)
          })

# we expect 6709291 to be included in both species lists
test_that("UPLOAD FAILED: 6709291 not included",
          {
            expect_true(6709291 %in% df_species)
            expect_true(6709291 %in% post_update_species)
          })
```

### Baseline

We get occurrence data of non alien species grouped at class level, which will be used for correcting the research bias effort:

```{r get_data_baseline}
df_bl <- read_csv("./cube_data/be_classes_cube.csv")
```

We add gemeente and provincie to the baseline data cube:

```{r}
df_bl <- df_bl %>% 
  left_join(utm1_gemeentes_provincies, by = c("eea_cell_code" = "CELLCODE")) %>% 
  st_drop_geometry() %>% 
  filter(!is.na(isFlanders)|!is.na(isWallonia)|!is.na(isBrussels))
```

```{r free up some ram 2}
rm(utm1_gemeentes_provincies)
gc() #garbage collection
```

Nota: these data have been also published on [Zenodo](https://doi.org/10.5281/zenodo.3632749).

We remove columns we will never use:

```{r remove_coords_uncertainty_baseline}
df_bl <-
  df_bl %>%
  select(-min_coord_uncertainty) %>% 
  filter(!is.na(classKey))
```

## Checklist data

We read input data based on [Global Register of Introduced and Invasive Species - Belgium](https://www.gbif.org/dataset/6d9e952f-948c-4483-9807-575348147c7e), typically referred to as the _unified checklist_. These data contain taxonomic information as well distribution, description and species profiles. We limit to taxa introduced in Belgium (`locationId` : `ISO_3166:BE`)

The input data is update monthly by the github action `get_griis_checklist.yaml`

```{r unified_checklist_tidy_df}
data_file <- here::here(
  "data",
  "output",
  "UAT_processing",
  "data_input_checklist_indicators.tsv"
)

taxa_df <-
  read_tsv(data_file,
           na = "",
           guess_max = 5000
  )

taxa_df <-
  taxa_df %>%
  filter(locationId == "ISO_3166:BE")
```

## Remove some taxa

### Taxa in unified checklist without occurrences

There is a group of taxa defined in the unified checklist which are not present in the occurrence data, i.e. the presence of these taxa, although confirmed by authoritative's checklists, is still not sustained by observations:

```{r taxa_without_occs}
alien_taxa_without_occs <-
  taxa_df %>%
  filter(!nubKey %in% df$taxonKey) %>%
  distinct(
    key,
    nubKey,
    canonicalName,
    first_observed,
    last_observed,
    class,
    kingdom
  ) %>%
  arrange(nubKey)
alien_taxa_without_occs
```

where `key` is the unique GBIF identifier of the taxon as published in the unified checklist, while `nubKey` is the unique GBIF identifier of the taxon in the GBIF Backbone Taxonomy.
We save this data.frame as `alien_taxa_without_occurrences.tsv` in `data/interim`:

```{r save_alien_taxa_without_occs}
write_tsv(
  alien_taxa_without_occs,
  here::here("data", "interim", "alien_taxa_without_occs.tsv"),
  na = ""
)
```

## Grid cells intersecting protected areas

We get information about intersection of grid cells with Belgian protected areas by reading file `intersect_EEA_ref_grid_protected_areas.tsv` in `data/output`:

```{r read_protected_areas}
df_prot_areas <- read_tsv(
  here::here(
    "data",
    "interim",
    "intersect_EEA_ref_grid_protected_areas.tsv"
  ),
  na = ""
)
```

Preview:

```{r preview_prot_areas}
df_prot_areas %>%
  head()
```

We are interested to areas included in Natura2000. We remove columns we are not interested to:

```{r select_cols_prot_areas}
df_prot_areas <-
  df_prot_areas %>%
  select(
    CELLCODE,
    natura2000
  )
```

#  Pre-processing

## Add informative columns

For better interpretation of the data, we retrieve scientific names (without authorship) of the species whose data we are going to analyze. We also retrieve the class of each alien species to link occurrence data with the related baseline data:

```{r get_canonical_names}
taxon_key <-
  df %>%
  distinct(taxonKey) %>%
  pull()
# get basic species information
pb <- progress_bar$new(total = length(taxon_key))
spec_names_raw <- map_df(
  taxon_key,
  function(k) {
    pb$tick()
    name_usage(key = k)$data
  }
) 

spec_names <- spec_names_raw %>%
  mutate(scientificName = coalesce(species, genus, family)) %>% 
  select(
    taxonKey = key,
    canonicalName,
    scientificName,
    rank,
    taxonomicStatus,
    kingdomKey, 
    classKey,
    authorship
  ) %>%
  mutate(canonicalName = ifelse(
    is.na(canonicalName), scientificName, canonicalName
  ))
```

```{r test spec_names}
taxa_list <- c(unique(spec_names$taxonKey), unique(alien_taxa_without_occs$nubKey))

# Test whether all taxa of the GRIIS checklist are included either in de occurrences cube or the list of species without occurrences
test_that("not all species included",
          {
            expect_in(unique(data_input_checklist_indicators$nubKey), taxa_list)
          })

# Test whether all taxa are unique
test_that("duplicate species",
          {
            expect_equal(length(unique(spec_names$taxonKey)), length(spec_names$taxonKey))
          })

# test whether all taxa have a scientific name
test_that("missing scientific names",
          {
            expect_false(any(is.na(spec_names$scientificName)))
          })

# test whether all taxa have a taxonKey
test_that("missing taxonKey",
          {
            expect_false(any(is.na(spec_names$taxonKey)))
          })
```

In the rare case multiple taxa share the same scientific name, authorship is added:

```{r add_authorship_if_needed}
spec_names <-
  spec_names %>%
  group_by(canonicalName) %>%
  add_tally() %>%
  ungroup() %>%
  mutate(canonicalName = if_else(n > 1,
                                 scientificName,
                                 canonicalName)) %>%
  select(-c(n))
```

We also retrieve kingdom and class:

```{r get_kingdom_class}
class_key <-
  spec_names %>%
  distinct(classKey) %>%
  filter(!is.na(classKey)) %>%
  pull()
pb <- progress_bar$new(total = length(class_key))
kingdom_class <- map_df(
  class_key,
  function(x) {
    pb$tick()
    name_usage(key = x)$data
  }
) %>%
  select(classKey, class, kingdomKey, kingdom)
```

and we add them:

```{r add_kingdom_class}
# add class
spec_names <-
  spec_names %>%
  left_join(kingdom_class %>%
              distinct(.data$classKey, .data$class),
            by = "classKey"
  )

# add kingdom
spec_names <-
  spec_names %>%
  left_join(kingdom_class %>%
              distinct(.data$kingdomKey, .data$kingdom),
            by = "kingdomKey"
  ) 
```

Preview:

```{r preview_species}
spec_names %>% head(10)
```

Taxa not assigned to any class:

```{r taxa_no_class}
spec_names %>%
  filter(is.na(classKey))
```

```{r complete keys}
keys_old <- alienSpecies::loadMetaData(type = "keys")

# backup current keys.csv file
keys_old %>%
  write_csv(here("data", "interim", "keys_old.csv"))

# remove columns we will not use
keys_new <- spec_names %>%
  select(-canonicalName, -authorship)

# rename columns to avoid confusion
colnames(keys_new) <- paste0(colnames(keys_new), "_new")

# Complete knowns, info in cube on checklist
keys_completed <- keys_old %>%
  full_join(keys_new, by = c("scientificName" = "scientificName_new")) %>% 
  mutate(taxonKey = coalesce(taxonKey, taxonKey_new),
         rank = coalesce(rank, rank_new),
         taxonomicStatus = coalesce(taxonomicStatus, taxonomicStatus_new),
         classKey = coalesce(classKey, classKey_new),
         kingdom = coalesce(kingdom, kingdom_new)) %>% 
  select(-contains("new")) %>%
  distinct()

# Find rows with missing taxonKey
keys_missing_taxonkey <- keys_completed %>% 
  filter(is.na(taxonKey))

# Function to derive the most likely taxonKey for a given scientific name
get_derived_taxonKey <- function(scientific_name) {
  taxonomic_data <- tryCatch(
    rgbif::name_usage(name = scientific_name)$data,
    error = function(e) NULL
  )
  if (is.null(taxonomic_data) || nrow(taxonomic_data) == 0) return(NA_integer_)
  
  derived_taxonKey <- taxonomic_data %>% 
    dplyr::filter(taxonomicStatus == "ACCEPTED", !is.na(nubKey)) %>% 
    group_by(nubKey) %>% 
    add_tally() %>% 
    distinct(nubKey, n) %>% 
    ungroup() %>% 
    filter(n == max(n, na.rm = TRUE)) %>% 
    pull(nubKey)
  
  # If multiple, take the first
  if (length(derived_taxonKey) > 0) return(derived_taxonKey[1])
  else return(NA_integer_)
}

# Use lapply to process all missing names
derived_keys <- pbapply::pblapply(keys_missing_taxonkey$scientificName, get_derived_taxonKey)
names(derived_keys) <- keys_missing_taxonkey$scientificName

# Update keys_completed with the derived taxonKeys
keys_completed <- keys_completed %>%
  mutate(
    taxonKey = ifelse(
      is.na(taxonKey) & scientificName %in% names(derived_keys),
      unlist(derived_keys[scientificName]),
      taxonKey
    )
  )

keys_path <- here("data", "output", "UAT_direct", "keys.csv")

write.csv(keys_completed, keys_path, row.names = FALSE)

# Upload to S3 bucket
aws.s3::put_object(file = keys_path,
                   object = "keys.csv",
                   bucket = UAT_bucket,
                   multipart = TRUE,
                   show_progress = TRUE,
                   region = "eu-west-1") 


# Complete unknowns, info not on cube or checklist
## Missing taxonkeys 
# => might not be an issue? => monitoring for now
# Fixing this will be a pain in the ass since `name_usage` returns every use of the scientific name in the gbif backbone as well as on very checklist worldwide
keys_missing_taxonkey <- keys_completed %>% 
  filter(is.na(taxonKey)) %>% 
  write_csv(here("data", "interim", "keys_missing_taxonkey.csv"))

## Missing scientific names => cause of https://github.com/inbo/alien-species-portal/issues/140
keys_missing_name <- keys_completed %>% 
  filter(is.na(scientificName)) %>% 
  write_csv(here("data", "interim", "keys_missing_name.csv"))

# test whether any missing scientificnames remain
test_that("missing scientificNames",
          {
            expect_false(nrow(keys_missing_name) > 0)
          })
```

## Extract x-y coordinates from cellcodes

Extract coordinates from column `eea_cell_code`. For example, the coordinates of cellcode `1kmE3895N3116` are `x = 3895` and `y = 3116`:

```{r extract_x_y}
df_xy <-
  df %>%
  distinct(eea_cell_code) %>%
  bind_cols(
    tibble(
      x = unlist(str_extract_all(unique(df$eea_cell_code),
                                 pattern = "(?<=E)\\d+"
      )),
      y = unlist(str_extract_all(unique(df$eea_cell_code),
                                 pattern = "(?<=N)\\d+"
      ))
    ) %>%
      mutate_all(as.integer)
  )
```

```{r extract_x_y - baseline}
df_bl_xy <-
  df_bl %>%
  tidylog::distinct(eea_cell_code) %>%
  bind_cols(
    tibble(
      x = unlist(str_extract_all(unique(df_bl$eea_cell_code),
                                 pattern = "(?<=E)[0-9\\-]+"
      )),
      y = unlist(str_extract_all(unique(df_bl$eea_cell_code),
                                 pattern = "(?<=N)[0-9\\-]+"
      ))
    ) %>%
      mutate_all(as.integer)
  )
```

Preview:

```{r x_y_preview, eval=FALSE}
df_xy %>% head()
df_bl_xy %>% head()
```

## Apply temporal and spatial restraints

Recalculate last observed date for each taxon:
using last observation date from the occurrence data, if available, or the last observation date from the unified checklist, if available.

```{r recalculate_last_observed_date, eval=FALSE}
taxa_df <- 
  taxa_df %>% 
  mutate(last_observed_old = as.integer(last_observed)) %>% 
  select(-last_observed) %>% 
  left_join(last_observed_taxon %>% 
              mutate(last_observed_new = as.integer(last_observed)) %>% 
              select(-last_observed), 
            by = c("nubKey" = "taxonKey")) %>% 
  mutate(last_observed = case_when(last_observed_old == last_observed_new ~ last_observed_old,
                                   last_observed_old < last_observed_new ~ last_observed_new,
                                   is.na(last_observed_old) ~ last_observed_new,
                                   is.na(last_observed_new) ~ last_observed_old,
                                   last_observed_new < last_observed_old ~ last_observed_old,
                                   TRUE ~ NA_integer_)) %>% 
  select(-last_observed_old, -last_observed_new)
```

Select alien species that were observed after 1950 (`year_cutoff`: 1950) or whose date of introduction is not known:

```{r select_aliens_with_recent_observations}
year_cutoff <- 1950

recent_alien_species <-
  taxa_df %>%
  # lower limit on date of first introduction if present
  filter(last_observed >= year_cutoff | is.na(last_observed)) %>%
  # remove duplicates due to other columns as pathways
  distinct(nubKey, first_observed, last_observed) %>%
  
  # get classkey info and filter out taxa without data
  inner_join(spec_names, by = c("nubKey" = "taxonKey")) %>%
  
  distinct(
    taxonKey = nubKey,
    .data$canonicalName,
    .data$first_observed,
    .data$last_observed,
    .data$class,
    .data$kingdom,
    .data$classKey,
    .data$kingdomKey
  )
```

Number of taxa last observed since 1950 or whose introduction year is not specified:

```{r n_recent_taxa}
nrow(recent_alien_species)
```

List of alien species excluded as last observation date is earlier than 1950:

```{r extinct_alien_species}
old_introductions_taxa <-
  taxa_df %>%
  filter(last_observed < year_cutoff) %>%
  anti_join(recent_alien_species %>%
              select(taxonKey),
            by = c("nubKey" = "taxonKey")
  ) %>%
  distinct(
    nubKey,
    first_observed,
    last_observed
  ) %>%
  inner_join(spec_names, by = c("nubKey" = "taxonKey")) %>%
  select(
    taxonKey = nubKey,
    canonicalName,
    first_observed,
    last_observed,
    class,
    kingdom,
    everything()
  )

old_introductions_taxa
```

We save them in file `taxa_last_observed_in_BE_before_1950.tsv` in  `data/output`:

```{r save_taxa_last_observed_before_1950_as_tsv_file}
write_tsv(old_introductions_taxa,
          here::here(
            "data",
            "interim",
            paste0(
              "taxa_last_observed_in_BE_before_",
              year_cutoff,
              ".tsv"
            )
          ),
          na = ""
)
```

We also limit the time series analysis to species which were last observed in 1950 or later, and limit their observations to after 1950, as this is the start of state of invasion researches

```{r define_year_range}
first_year <- year_cutoff
```

Some observations can be published with erroneous coordinates (see [#trias-project/occ-processing#13](https://github.com/trias-project/occ-processing/issues/13)) or can be assigned to a grid cell out of Belgium and should therefore removed.

```{r define_x_y_range_values}
be_cellcodes <- df_prot_areas$CELLCODE
```

Apply spatial and temporal restraints defined above to both occurrence data:

```{r apply_restraints_time_space_occ_data}
df <-
  df %>%
  # recently introduced taxa
  filter(taxonKey %in% recent_alien_species$taxonKey) %>%
  left_join(df_xy, by = "eea_cell_code") %>%
  # recent history
  filter(year > first_year) %>%
  # in Belgium
  filter(eea_cell_code %in% be_cellcodes)
```

```{r add restraints to baseline data}
df_bl <-
  df_bl %>%
  tidylog::left_join(df_bl_xy, by = "eea_cell_code") %>%
  # Recent history
  tidylog::filter(year > first_year) %>% 
  # in Belgium
  tidylog::filter(eea_cell_code %in% be_cellcodes)
```

## Create time series

For each species, define cells with at least one observation:

```{r distinct_cells_taxa}
df_cc <- 
  df %>%
  group_by(taxonKey) %>%
  distinct(eea_cell_code) %>%
  ungroup()
```

For each species, identify the first year with at least one observation:

```{r begin_year_per_species}
df_begin_year <- 
  df %>%
  group_by(taxonKey) %>%
  summarize(begin_year = min(year))
```

For each species, combine `begin_year` and unique `eea_cell_code` as found above: 

```{r combine_begin_year_cells_taxa}
df_cc <- 
  df_cc %>%
  left_join(df_begin_year, by = "taxonKey") %>%
  select(taxonKey, begin_year, eea_cell_code)
```

Preview:

```{r preview_begin_year_cells_combo}
df_cc %>% head()
```

For each cell (`eea_cell_code`) and species (`taxonKey`) we can now create a time series:

```{r fun: make_ts, cache=TRUE}
make_time_series <- function(eea_cell_code, taxonKey, begin_year, last_year ) {
  expand_grid(eea_cell_code = eea_cell_code,
              taxonKey = taxonKey,
              year = seq(from = begin_year, to = last_year))
  
}
```

```{r create_timeseries}
# create timeseries slots
df_ts <- pmap_dfr(df_cc, 
                  .f = make_time_series, 
                  last_year = year(Sys.Date())
)
```

```{r cleanup df_cc, message=NA}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df_cc to free up some memory")
  remove(df_cc)
  gc()
}
```  

## Add data

# add occurrence data

```{r add_occurrence_data}
df <- df %>% 
  select(taxonKey, year, eea_cell_code, obs)

gc()

df_ts <- 
  df_ts %>%
  left_join(df, 
            by = c("taxonKey", "year", "eea_cell_code"))
```

# remove df to free up some memory

```{r remove_df}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df to free up some memory")
  remove(df)
  gc()
}
```

# add membership to protected areas

```{r add_protected_areas}
df_prot_areas <- df_prot_areas %>% 
  select(CELLCODE, natura2000)

gc()

df_ts <- 
  df_ts %>%
  left_join(df_prot_areas,
            by = c("eea_cell_code" = "CELLCODE"))
```

```{r remove_prot_areas}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df_prot_areas to free up some memory")
  remove(df_prot_areas)
  gc()
}
```  

# add classKey

```{r add_classKey}
df_ts <- 
  df_ts %>%
  left_join(spec_names %>% 
              select(taxonKey, classKey), 
            by = "taxonKey")
```

## Research effort correction

```{r Add baseline data}
# at class level diminished by obs of specific alien taxon
df_bl <- df_bl %>% 
  select(year, eea_cell_code, classKey, cobs) %>% 
  distinct()

gc()

nrow_prior <- nrow(df_ts)

df_ts <-
  df_ts %>%
  left_join(
    df_bl,
    by = c("year", "eea_cell_code", "classKey")
  )

if(nrow_prior != nrow(df_ts)){
  stop("Number of rows changed after join")
}
```

```{r remove_df_bl}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df_bl to free up some memory")
  remove(df_bl)
  gc()
}
```

To correct the effect of research effort of an alien speies, we calculate the number of observations at class level excluding the observations of the alien species itself:

```{r correct_cobs_research_effort_bias}
# add baseline data (at class level) diminished by obs of specific alien taxon
df_ts <- 
  df_ts %>%
  mutate(cobs = cobs - obs)

df_ts_below_zero <- df_ts %>% filter(cobs < 0)

if(nrow(df_ts_below_zero) > 0){
  warning("Negative values in cobs column")
  
  # replace negative values with 0
  df_ts <- 
    df_ts %>%
    mutate(cobs = if_else(cobs < 0, 0, cobs))
}

# replace NAs with 0
df_ts <-
  df_ts %>%
  replace_na(list(cobs = 0, obs = 0))
```

Preview:

```{r preview_df_ts}
df_ts %>% head(n = 30)
```

Add column for presence (1) or absence (0)

```{r add_presence_absence}
df_ts <- 
  df_ts %>%
  mutate(pa_cobs = if_else(cobs > 0, 1, 0),
         pa_obs = if_else(obs > 0, 1, 0))
```

Arrange order columns:

```{r set_order_columns}
df_ts <-
  df_ts %>%
  select(taxonKey, 
         year, 
         eea_cell_code, 
         obs, 
         pa_obs, 
         cobs, 
         pa_cobs,
         classKey,
         natura2000)
```

# Save data

## Save timeseries and taxonomic metadata

Save `df_ts` as it will be used as start point of modelling pipelines:

```{r save_data_as_tsv}
# upload data to S3 bucket
aws.s3::s3save(df_ts, 
               object = "df_timeseries.Rdata", 
               bucket = UAT_bucket, 
               opts = list(show_progress = TRUE,
                           multipart = TRUE))
```

Create time series data
input:  "df_timeseries.RData" and "grid.RData" from bucket
output: full_timeseries.RData

```{r create timeseries}
alienSpecies::createTimeseries(
  # read grid.RData from bucket
  shapeData = alienSpecies::loadShapeData("grid.RData")$utm1_bel_with_regions, 
  bucket = UAT_bucket
)
```

reload previously loaded inbotheme version
```{r reload_inbotheme_version, eval=TRUE}
if(Sys.getenv("amiabot") != "yes"){
  unloadNamespace("alienSpecies")
  if(loaded_inbotheme_version != "0.5.8"){
    remove.packages("INBOtheme")
    devtools::install_github(paste0("inbo/INBOtheme@", loaded_inbotheme_version))
  }
}
```
