---
title: "update_trendOccupancy"
author: "Sander Devisscher"
date: "2025-07-30"
output: html_document
---

Trend occupancy or the evolution of union concern species according to the 
National Scientific Secretariat on Invasive Alien Species (NSSIAS) to the European Union 
via the EASIN reporting tool. 
This evolution consists of the number of 10x10km squares occupied by a species at 
the moment of inclusion onto the list (t0) compared to the status at the moment of 
a 6y reporting cycle (tx).

```{r libraries}
library(dplyr)
library(magrittr)
library(testthat)
library(sf)
library(readr)
library(stringr)
library(tidyr)
```

```{r download zenodo files}
t0_url <- "https://zenodo.org/api/records/7708520/files-archive"
t0_files <- c("ias_belgium_t0_2016.geojson", 
              "ias_belgium_t0_2018.geojson", 
              "ias_belgium_t0_2020.geojson", 
              "ias_belgium_t0_2023.geojson")
t1_url <- "https://zenodo.org/api/records/3060173/files-archive"
t1_files <- c("T1_Belgium_Union_List_Species.shp")
t2_url <- "" #Placeholder future t2 url. Also add 't2' to 'tx' & add a period for 't2'

tx <- c("t0", "t1")

for(x in tx){
  destfile <- paste0(tempdir(), "/", x, ".zip")
  url <- get(paste0(x, "_url"))
  tx_files <- get(paste0(x, "_files"))
  exdir <- paste0(tempdir(), "/", x, "/")
  
  download.file(url, destfile, mode = "wb")
  unzip(destfile, exdir = exdir)
  
  tx_file <- data.frame()
  
  for(f in tx_files){
    temp <- sf::st_read(paste0(exdir, f))
    
    # Determine period
    if(x == "t0"){
      period  <- paste("2000-", str_extract(f, "\\d{4}"))
    }else{
      if(x == "t1"){
        period <- "2015-2018"
      }
      if(!x %in% c("t0", "t1")){
        stop(paste("Set the period for", x))
      }
    }
    
    # add period
    temp$period <- period
    
    if("REFERENCE_" %in% colnames(temp)){
      temp <- temp %>% 
        rename("Reference" = "REFERENCE_")
    }
    
    if("REFERENCE" %in% colnames(temp)){
      temp <- temp %>% 
        rename("Reference" = "REFERENCE")
    }
    
    if("DATA_PARTN" %in% colnames(temp)){
      temp <- temp %>% 
        rename("Data_Partn" = "DATA_PARTN")
    }
    
    if("CellCode" %in% colnames(temp)){
       temp <- temp %>% 
        rename("Cellcode" = "CellCode")
    }
    
    if(nrow(tx_file) == 0){
      tx_file <- temp
    }else{
      tx_file <- rbind(tx_file, temp)
    }
  }
  
  if("species" %in% colnames(tx_file)){
    tx_file <- tx_file %>% 
      rename("Species" = "species")
  }
  
  tx_file$source <- x
  
  assign(paste0(x, "_raw"), tx_file)
}

table(t0_raw$period)
```

```{r set creds}
Sys.setenv("AWS_DEFAULT_REGION" = "eu-west-1")

if(Sys.getenv("S3_BUCKET") != ""){
  UAT_bucket <- Sys.getenv("S3_BUCKET")
}else{
  UAT_bucket <- Sys.getenv("UAT_bucket")
}
```

```{r connect to the bucket}
if(Sys.getenv("amiabot") != "yes"){
  print("Executor of the script is a human >> connecting to bucket")
  source("./src/connect_to_bucket.R")
  connect_to_bucket(bucket_name = UAT_bucket)
}
```

```{r get keys}
keys <- alienSpecies::loadMetaData(type = "keys")
```

```{r T0: check & correct species}
table(t0_raw$Species, useNA = "ifany")

t0_cleaned <- t0_raw %>% 
  filter(!is.na(Species)) %>% 
  mutate(Species = case_match(Species, 
                              "Orconectes limosus" ~ "Faxonius limosus",
                              .default = Species)) 

t0_species_not_keys <- t0_cleaned %>% 
  filter(Accepted != "N") %>% 
  filter(!Species %in% keys$scientificName) %>% 
  distinct(Species) %>% 
  pull()

t0_cleaned <- t0_cleaned %>% 
  as.data.frame() %>% 
  filter(Accepted != "N") %>% # Filter rejected squares
  select(species = Species, source, cell_code10 = Cellcode, period)
  

test_that("t0_species_not_keys should have length 0", {
  if (length(t0_species_not_keys) != 0) {
    print(t0_species_not_keys)
  }
  expect_length(t0_species_not_keys, 0)
})
```

```{r T1: check & correct species}
table(t1_raw$Species, useNA = "ifany")

t1_cleaned <- t1_raw %>% 
  filter(!is.na(Species)) %>% 
  mutate(Species = gsub("\\s*\\([^\\)]+\\)", "", Species)) %>% 
  separate(col = "Species",
           into = c("Species_1", "Species_2"),
           sep = " ",
           remove = TRUE,
           extra = "drop") %>% 
  mutate(Species = paste(Species_1, Species_2)) %>% 
  mutate(Species = case_match(Species, 
                              "Orconectes limosus" ~ "Faxonius limosus",
                              "Trachemys Agassiz," ~ "Trachemys scripta",
                              .default = Species)) 

table(t1_cleaned$Species, useNA = "ifany")

t1_species_not_keys <- t1_cleaned %>% 
  filter(!Species %in% keys$scientificName) %>% 
  distinct(Species) %>% 
  pull()

t1_cleaned <- t1_cleaned %>% 
  as.data.frame() %>% 
  select(species = Species, source, cell_code10 = Cellcode, period)

test_that("t1_species_not_keys should have length 0", {
  if (length(t1_species_not_keys) != 0) {
    print(t1_species_not_keys)
  }
  expect_length(t1_species_not_keys, 0)
})
```

```{r combine tx}
trendOccupancy_belgium <- rbind(t0_cleaned, t1_cleaned)

table(trendOccupancy_belgium$species, trendOccupancy_belgium$period, useNA = "ifany")
```

```{r export files}
write_csv(trendOccupancy_belgium, "./data/output/UAT_processing/trendOccupancy/trendOccupancy_belgium.csv")
```

