---
title: "create_timeseries_decade"
author: "Sander Devisscher"
date: "2026-03-11"
output: html_document
params:
  decade_min: 2010
---

```{r load_libraries, message = FALSE}
library(dplyr) # To do data science
library(readr)
library(magrittr)
library(purrr)
library(stringr)
library(tidyr)
library(tidylog) # To provide feedback on dplyr functions
library(progress) # To add progress bars
library(here) # To find files
library(lubridate) # To work with dates
library(rgbif) # To get taxa from publisehd unified checklist
library(sf) # To work with spatial data
library(testthat) # To test functions
```

```{r load correct version of INBOtheme, eval=TRUE}
if(Sys.getenv("amiabot") != "yes"){
  print("Executor of the script is a human >> checking INBOtheme version")
  # determine the loaded INBOtheme version
  loaded_inbotheme <- installed.packages() %>% 
    as_tibble() %>% 
    filter(Package == "INBOtheme")
  loaded_inbotheme_version <- loaded_inbotheme$Version
  
  if(is_empty(loaded_inbotheme_version)){
    print("INBOtheme is not installed >> installing INBOtheme")
    devtools::install_github("inbo/INBOtheme@v0.5.8")
  }
  
  if(length(loaded_inbotheme_version) > 1 || loaded_inbotheme_version != "0.5.8"){
    print("INBOtheme version is not correct >> loading correct version")
    # remove installed INBOtheme
    remove.packages("INBOtheme")
    
    # load correct version of INBOtheme
    devtools::install_github("inbo/INBOtheme@v0.5.8")
  }else{
    print("Correct INBOtheme version is loaded")
  }
}else{
  print("Executor of the script is a bot >> correct INBOtheme version is loaded")
}

library(INBOtheme)
```

# Connect to bucket
## set credentials
```{r set creds}
Sys.setenv("AWS_DEFAULT_REGION" = "eu-west-1")

if(Sys.getenv("S3_BUCKET") != ""){
  UAT_bucket <- Sys.getenv("S3_BUCKET")
}else{
  UAT_bucket <- Sys.getenv("UAT_bucket")
}
```

## connect to bucket
```{r connect to bucket}
if(Sys.getenv("amiabot") != "yes"){
  print("Executor of the script is a human >> connecting to bucket")
  source("./src/connect_to_bucket.R")
  connect_to_bucket(bucket_name = UAT_bucket)
}else{
  test_that("AWS credentials are set in the environment", {
    # 1. Check Access Key
    expect_true(Sys.getenv("AWS_ACCESS_KEY_ID") != "", 
                info = "AWS_ACCESS_KEY_ID is missing or empty")
    
    # 2. Check Secret Key
    expect_true(Sys.getenv("AWS_SECRET_ACCESS_KEY") != "", 
                info = "AWS_SECRET_ACCESS_KEY is missing or empty")
    
    # 3. Check Session Token (Required since you are using role-to-assume)
    expect_true(Sys.getenv("AWS_SESSION_TOKEN") != "", 
                info = "AWS_SESSION_TOKEN is missing (Required for assumed roles)")
    
  })
}
```

### Baseline

```{r set decade}
decade_min <- as.integer(params$decade_min)
decade_max <- decade_min + 9

# Overwrite decade_max when larger than current year
if(decade_max > year(Sys.Date())){
  decade_max <- year(Sys.Date())
}

cat("Calculating", decade_min, "-", decade_max, "timeframe \n")
```

```{r get_data_cube}
df <- read_csv("./cube_data/be_alientaxa_cube_processed.csv") %>% 
  filter(year >= decade_min,
         year <= decade_max)

table(df$year, useNA = "ifany")
```

We get occurrence data of non alien species grouped at class level, which will be used for correcting the research bias effort:

```{r get_data_baseline}
df_bl <- read_csv("./cube_data/be_classes_cube.csv") %>% 
  filter(year >= decade_min,
         year <= decade_max)

table(df_bl$year, useNA = "ifany")
```

We add gemeente and provincie to the baseline data cube:

```{r read utm1_gemeentes_provincies}
utm1_gemeentes_provincies <- sf::st_read("./data/output/UAT_processing/grid/utm1_bel_with_regions.gpkg", 
                                         layer="utm1_bel_with_regions" )
```

## Checklist data

We read input data based on [Global Register of Introduced and Invasive Species - Belgium](https://www.gbif.org/dataset/6d9e952f-948c-4483-9807-575348147c7e), typically referred to as the _unified checklist_. These data contain taxonomic information as well distribution, description and species profiles. We limit to taxa introduced in Belgium (`locationId` : `ISO_3166:BE`) and exclude species that where not present in Belgium during the defined decade.

```{r read checklist}
data_input_checklist_indicators <- read_tsv("./data/output/UAT_processing/data_input_checklist_indicators.tsv") %>% 
  rowwise() %>% 
  mutate(first_observed = replace_na(first_observed,
                                     1950),  
         last_observed = ifelse(first_observed > last_observed, 
                                first_observed, 
                                last_observed),
         last_observed = replace_na(last_observed,
                                    first_observed))

#first_observed -- decade_min -- decade_max -- last_observed
presence_wider_1 <- data_input_checklist_indicators %>% 
  filter(first_observed <= decade_min & last_observed >= decade_max)
#decade_min -- first_observed -- decade_max -- last_observed
presence_wider_2 <- data_input_checklist_indicators %>% 
  filter(!species %in% presence_wider_1$species) %>% 
  filter(first_observed >= decade_min & 
           first_observed <= decade_max & 
           last_observed >= decade_max)
#first_observed -- decade_min -- last_observed -- decade_max
presence_wider_3 <- data_input_checklist_indicators %>% 
  filter(!species %in% presence_wider_1$species) %>% 
  filter(!species %in% presence_wider_2$species) %>% 
  filter(first_observed <= decade_min & 
           (last_observed <= decade_max & 
              last_observed >= decade_min))

presence_wider <- rbind(presence_wider_1, presence_wider_2, presence_wider_3)

#decade_min -- first_observed -- last_observed -- decade_max
presence_within <- data_input_checklist_indicators %>% 
  filter(!species %in% presence_wider$species) %>% 
  filter(first_observed >= decade_min & last_observed <= decade_max)

spec_names <- read_csv("./data/interim/spec_names.csv")

taxa_df <- bind_rows(presence_wider, presence_within) %>%
  dplyr::filter(locationId == "ISO_3166:BE") %>% 
  select(-class, -classKey, -kingdom, -kingdomKey) %>% 
  left_join(spec_names, 
            by = c("nubKey" = "taxonKey")) 

table(taxa_df$first_observed, useNA = "ifany")
table(taxa_df$last_observed, useNA = "ifany")
```

```{r}
df_bl <- df_bl %>% 
  left_join(utm1_gemeentes_provincies, by = c("eea_cell_code" = "CELLCODE")) %>% 
  st_drop_geometry() %>% 
  filter(!is.na(isFlanders)|!is.na(isWallonia)|!is.na(isBrussels))
```

```{r free up some ram 2}
rm(utm1_gemeentes_provincies)
gc() #garbage collection
```

We remove columns we will never use:

```{r remove_coords_uncertainty_baseline}
df_bl <-
  df_bl %>%
  select(-min_coord_uncertainty) %>% 
  filter(!is.na(classKey))
```

## Remove some taxa

### Taxa in unified checklist without occurrences

There is a group of taxa defined in the unified checklist which are not present in the occurrence data, i.e. the presence of these taxa, although confirmed by authoritative's checklists, is still not sustained by observations:

```{r taxa_without_occs}
alien_taxa_without_occs <-
  taxa_df %>%
  filter(!nubKey %in% df$taxonKey) %>%
  distinct(
    key,
    nubKey,
    canonicalName,
    first_observed,
    last_observed,
    class,
    kingdom
  ) %>%
  arrange(nubKey)
alien_taxa_without_occs
```

where `key` is the unique GBIF identifier of the taxon as published in the unified checklist, while `nubKey` is the unique GBIF identifier of the taxon in the GBIF Backbone Taxonomy.
We save this data.frame as `alien_taxa_without_occurrences.tsv` in `data/interim`:

```{r save_alien_taxa_without_occs}
write_tsv(
  alien_taxa_without_occs,
  here::here("data", "interim", paste0("alien_taxa_without_occs_", 
                                       decade_min, "_", decade_max, ".csv")),
  na = ""
)
```

## Grid cells intersecting protected areas

We get information about intersection of grid cells with Belgian protected areas by reading file `intersect_EEA_ref_grid_protected_areas.tsv` in `data/output`:

```{r read_protected_areas}
df_prot_areas <- read_tsv(
  here::here(
    "data",
    "interim",
    "intersect_EEA_ref_grid_protected_areas.tsv"
  ),
  na = ""
)
```

Preview:

```{r preview_prot_areas}
df_prot_areas %>%
  head()
```

We are interested to areas included in Natura2000. We remove columns we are not interested to:

```{r select_cols_prot_areas}
df_prot_areas <-
  df_prot_areas %>%
  select(
    CELLCODE,
    natura2000
  )
```

#  Pre-processing

## Add informative columns

For better interpretation of the data, we retrieve scientific names (without authorship) of the species whose data we are going to analyze. We also retrieve the class of each alien species to link occurrence data with the related baseline data:

```{r}
keys_path <- here("data", "output", "UAT_processing", "be_alientaxa_info.csv")

spec_names <- read_delim(keys_path, ",", quote = "needed") %>% 
  filter(taxonKey %in% taxa_df$nubKey)
```

```{r test spec_names}
taxa_list <- c(unique(spec_names$taxonKey), unique(alien_taxa_without_occs$nubKey))

# Test whether all taxa of the GRIIS checklist are included either in de occurrences cube or the list of species without occurrences
# test_that("not all species included",
#           {
#             expect_in(unique(taxa_df$nubKey), taxa_list)
#           })

# Test whether all taxa are unique
test_that("duplicate species",
          {
            expect_equal(length(unique(spec_names$taxonKey)), length(unique(spec_names$taxonKey)))
          })

# test whether all taxa have a scientific name
test_that("missing scientific names",
          {
            expect_false(any(is.na(spec_names$scientificName)))
          })

# test whether all taxa have a taxonKey
test_that("missing taxonKey",
          {
            expect_false(any(is.na(spec_names$taxonKey)))
          })

spec_names <- taxa_df %>% 
  select(taxonKey = nubKey,
         scientificName,
         canonicalName,
         rank,
         taxonomicStatus,
         class,
         classKey,
         kingdomKey,
         kingdom)
```

## Extract x-y coordinates from cellcodes

Extract coordinates from column `eea_cell_code`. For example, the coordinates of cellcode `1kmE3895N3116` are `x = 3895` and `y = 3116`:

```{r extract_x_y}
df_xy <-
  df %>%
  distinct(eea_cell_code) %>%
  bind_cols(
    tibble(
      x = unlist(str_extract_all(unique(df$eea_cell_code),
                                 pattern = "(?<=E)\\d+"
      )),
      y = unlist(str_extract_all(unique(df$eea_cell_code),
                                 pattern = "(?<=N)\\d+"
      ))
    ) %>%
      mutate_all(as.integer)
  )
```

```{r extract_x_y - baseline}
df_bl_xy <-
  df_bl %>%
  tidylog::distinct(eea_cell_code) %>%
  bind_cols(
    tibble(
      x = unlist(str_extract_all(unique(df_bl$eea_cell_code),
                                 pattern = "(?<=E)[0-9\\-]+"
      )),
      y = unlist(str_extract_all(unique(df_bl$eea_cell_code),
                                 pattern = "(?<=N)[0-9\\-]+"
      ))
    ) %>%
      mutate_all(as.integer)
  )
```

Preview:

```{r x_y_preview, eval=FALSE}
df_xy %>% head()
df_bl_xy %>% head()
```

## Apply temporal and spatial restraints

Recalculate last observed date for each taxon:
using last observation date from the occurrence data, if available, or the last observation date from the unified checklist, if available.

Select alien species that were observed after 1950 (`year_cutoff`: 1950) or whose date of introduction is not known:

```{r select_aliens_with_recent_observations}
year_cutoff <- 1950

recent_alien_species <-
  taxa_df %>%
  # lower limit on date of first introduction if present
  filter(last_observed >= year_cutoff | is.na(last_observed)) %>%
  # remove duplicates due to other columns as pathways
  distinct(nubKey, first_observed, last_observed) %>%
  
  # get classkey info and filter out taxa without data
  inner_join(spec_names, by = c("nubKey" = "taxonKey")) %>%
  
  distinct(
    taxonKey = nubKey,
    .data$canonicalName,
    .data$first_observed,
    .data$last_observed,
    .data$class,
    .data$kingdom,
    .data$classKey,
    .data$kingdomKey
  )
```

Number of taxa last observed since 1950 or whose introduction year is not specified:

```{r n_recent_taxa}
nrow(recent_alien_species)
```

List of alien species excluded as last observation date is earlier than 1950:

```{r extinct_alien_species}
old_introductions_taxa <-
  taxa_df %>%
  filter(last_observed < year_cutoff) %>%
  anti_join(recent_alien_species %>%
              select(taxonKey),
            by = c("nubKey" = "taxonKey")
  ) %>%
  distinct(
    nubKey,
    first_observed,
    last_observed
  ) %>%
  inner_join(spec_names, by = c("nubKey" = "taxonKey")) %>%
  select(
    taxonKey = nubKey,
    canonicalName,
    first_observed,
    last_observed,
    class,
    kingdom,
    everything()
  )

old_introductions_taxa
```

We save them in file `taxa_last_observed_in_BE_before_1950.tsv` in  `data/output`:

```{r save_taxa_last_observed_before_1950_as_tsv_file}
write_tsv(old_introductions_taxa,
          here::here(
            "data",
            "interim",
            paste0(
              "taxa_last_observed_in_BE_before_",
              year_cutoff,
              ".tsv"
            )
          ),
          na = ""
)
```

We limit the time series analysis to the first year of the decade (decade_min)

```{r define_year_range}
first_year <- decade_min
```

Some observations can be published with erroneous coordinates (see [#trias-project/occ-processing#13](https://github.com/trias-project/occ-processing/issues/13)) or can be assigned to a grid cell out of Belgium and should therefore removed.

```{r define_x_y_range_values}
be_cellcodes <- df_prot_areas$CELLCODE
```

Apply spatial and temporal restraints defined above to both occurrence data:

```{r apply_restraints_time_space_occ_data}
df <-
  df %>%
  # recently introduced taxa
  filter(taxonKey %in% recent_alien_species$taxonKey) %>%
  left_join(df_xy, by = "eea_cell_code") %>%
  # recent history
  filter(year >= first_year) %>%
  # in Belgium
  filter(eea_cell_code %in% be_cellcodes)
```

```{r add restraints to baseline data}
df_bl <-
  df_bl %>%
  tidylog::left_join(df_bl_xy, by = "eea_cell_code") %>%
  # Recent history
  tidylog::filter(year >= first_year) %>% 
  # in Belgium
  tidylog::filter(eea_cell_code %in% be_cellcodes)
```

## Create time series

For each species, define cells with at least one observation:

```{r distinct_cells_taxa}
df_cc <- 
  df %>%
  group_by(taxonKey) %>%
  distinct(eea_cell_code) %>%
  ungroup()
```

For each species, identify the first year with at least one observation:

```{r begin_year_per_species}
df_begin_year <- 
  df %>%
  group_by(taxonKey) %>%
  summarize(begin_year = min(year))
```

For each species, combine `begin_year` and unique `eea_cell_code` as found above: 

```{r combine_begin_year_cells_taxa}
df_cc <- 
  df_cc %>%
  left_join(df_begin_year, by = "taxonKey") %>%
  select(taxonKey, begin_year, eea_cell_code)
```

Preview:

```{r preview_begin_year_cells_combo}
df_cc %>% head()
```

For each cell (`eea_cell_code`) and species (`taxonKey`) we can now create a time series:

```{r fun: make_ts, cache=TRUE}
make_time_series <- function(eea_cell_code, taxonKey, begin_year, last_year ) {
  expand_grid(eea_cell_code = eea_cell_code,
              taxonKey = taxonKey,
              year = seq(from = begin_year, to = last_year))
}
```

```{r create_timeseries}
# create timeseries slots
df_ts <- pmap_dfr(df_cc, 
                  .f = make_time_series, 
                  last_year = decade_max)
```

```{r cleanup df_cc, message=NA}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df_cc to free up some memory")
  remove(df_cc)
  gc()
}
```  

## Add data

# add occurrence data

```{r add_occurrence_data}
df <- df %>% 
  select(taxonKey, year, eea_cell_code, obs)

gc()

df_ts <- 
  df_ts %>%
  left_join(df, 
            by = c("taxonKey", "year", "eea_cell_code"))
```

# remove df to free up some memory

```{r remove_df}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df to free up some memory")
  remove(df)
  gc()
}
```

# add membership to protected areas

```{r add_protected_areas}
df_prot_areas <- df_prot_areas %>% 
  select(CELLCODE, natura2000)

gc()

df_ts <- 
  df_ts %>%
  left_join(df_prot_areas,
            by = c("eea_cell_code" = "CELLCODE"))
```

```{r remove_prot_areas}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df_prot_areas to free up some memory")
  remove(df_prot_areas)
  gc()
}
```  

# add classKey

```{r add_classKey}
df_ts <- 
  df_ts %>%
  left_join(spec_names %>% 
              distinct(taxonKey, classKey), 
            by = "taxonKey")
```

## Research effort correction

```{r Add baseline data}
# at class level diminished by obs of specific alien taxon
df_bl <- df_bl %>% 
  select(year, eea_cell_code, classKey, cobs) %>% 
  distinct()

gc()

nrow_prior <- nrow(df_ts)

df_ts <-
  df_ts %>%
  left_join(
    df_bl,
    by = c("year", "eea_cell_code", "classKey")
  )

if(nrow_prior != nrow(df_ts)){
  stop("Number of rows changed after join")
}
```

```{r remove_df_bl}
if(Sys.getenv("amiabot") == "yes"){
  cat("Removing df_bl to free up some memory")
  remove(df_bl)
  gc()
}
```

To correct the effect of research effort of an alien speies, we calculate the number of observations at class level excluding the observations of the alien species itself:

```{r correct_cobs_research_effort_bias}
# add baseline data (at class level) diminished by obs of specific alien taxon
df_ts <- 
  df_ts %>%
  mutate(cobs = cobs - obs)

df_ts_below_zero <- df_ts %>% filter(cobs < 0)

if(nrow(df_ts_below_zero) > 0){
  warning("Negative values in cobs column")
  
  # replace negative values with 0
  df_ts <- 
    df_ts %>%
    mutate(cobs = if_else(cobs < 0, 0, cobs))
}

# replace NAs with 0
df_ts <-
  df_ts %>%
  replace_na(list(cobs = 0, obs = 0))
```

Preview:

```{r preview_df_ts}
df_ts %>% head(n = 30)
```

Add column for presence (1) or absence (0)

```{r add_presence_absence}
df_ts <- 
  df_ts %>%
  mutate(pa_cobs = if_else(cobs > 0, 1, 0),
         pa_obs = if_else(obs > 0, 1, 0))
```

Arrange order columns:

```{r set_order_columns}
df_ts <-
  df_ts %>%
  select(taxonKey, 
         year, 
         eea_cell_code, 
         obs, 
         pa_obs, 
         cobs, 
         pa_cobs,
         classKey,
         natura2000)
```

```{r export for artifact upload}
write_csv(df_ts, paste0("./cube_data/timeseries_", decade_min, "_", decade_max,".csv"))
```