---
title: "cube_processing"
author: "Sander Devisscher"
date: "2026-03-11"
output: html_document
---

```{r load_libraries, message = FALSE}
library(dplyr) # To do data science
library(readr)
library(magrittr)
library(purrr)
library(stringr)
library(tidyr)
library(tidylog) # To provide feedback on dplyr functions
library(progress) # To add progress bars
library(here) # To find files
library(lubridate) # To work with dates
library(rgbif) # To get taxa from publisehd unified checklist
library(sf) # To work with spatial data
library(testthat) # To test functions
```

```{r load correct version of INBOtheme, eval=TRUE}
if(Sys.getenv("amiabot") != "yes"){
  print("Executor of the script is a human >> checking INBOtheme version")
  # determine the loaded INBOtheme version
  loaded_inbotheme <- installed.packages() %>% 
    as_tibble() %>% 
    filter(Package == "INBOtheme")
  loaded_inbotheme_version <- loaded_inbotheme$Version
  
  if(is_empty(loaded_inbotheme_version)){
    print("INBOtheme is not installed >> installing INBOtheme")
    devtools::install_github("inbo/INBOtheme@v0.5.8")
  }
  
  if(length(loaded_inbotheme_version) > 1 || loaded_inbotheme_version != "0.5.8"){
    print("INBOtheme version is not correct >> loading correct version")
    # remove installed INBOtheme
    remove.packages("INBOtheme")
    
    # load correct version of INBOtheme
    devtools::install_github("inbo/INBOtheme@v0.5.8")
  }else{
    print("Correct INBOtheme version is loaded")
  }
}else{
  print("Executor of the script is a bot >> correct INBOtheme version is loaded")
}

library(INBOtheme)
```

# Connect to bucket
## set credentials
```{r set creds}
Sys.setenv("AWS_DEFAULT_REGION" = "eu-west-1")

if(Sys.getenv("S3_BUCKET") != ""){
  UAT_bucket <- Sys.getenv("S3_BUCKET")
}else{
  UAT_bucket <- Sys.getenv("UAT_bucket")
}
```

## connect to bucket
```{r connect to bucket}
if(Sys.getenv("amiabot") != "yes"){
  print("Executor of the script is a human >> connecting to bucket")
  source("./src/connect_to_bucket.R")
  connect_to_bucket(bucket_name = UAT_bucket)
}else{
  test_that("AWS credentials are set in the environment", {
    # 1. Check Access Key
    expect_true(Sys.getenv("AWS_ACCESS_KEY_ID") != "", 
                info = "AWS_ACCESS_KEY_ID is missing or empty")
    
    # 2. Check Secret Key
    expect_true(Sys.getenv("AWS_SECRET_ACCESS_KEY") != "", 
                info = "AWS_SECRET_ACCESS_KEY is missing or empty")
    
    # 3. Check Session Token (Required since you are using role-to-assume)
    expect_true(Sys.getenv("AWS_SESSION_TOKEN") != "", 
                info = "AWS_SESSION_TOKEN is missing (Required for assumed roles)")
    
  })
}
```

# Get data

## read utm1_bel_with_regions file
```{r read utm1_gemeentes_provincies}
utm1_gemeentes_provincies <- sf::st_read("./data/output/UAT_processing/grid/utm1_bel_with_regions.gpkg", 
                                         layer="utm1_bel_with_regions" )
```

We get occurrence data as obtained in `trias-project/occ-processing` repository. 

## Occurrence data

### Alien species

We load occurrence data related to alien species in Belgium:

```{r read_data_cube}
df <- read_csv("./cube_data/be_alientaxa_cube.csv") 

print(file.info("./cube_data/be_alientaxa_cube.csv"))
```

To save the data cube to the S3 bucket, we use the following code:
A temporary directory is created to store the data cube.
```{r create_temp_dir}
cube_path <- tempdir()
dir.create(cube_path, showWarnings = FALSE)
```

For display on invasion history maps of the alienspecies portal we add the 
gemeente and provincie to the data cube.
This is done by joining the data cube with the `utm1_gemeentes_provincies` dataset.
which is updated by the following code:
[processing_utm10_files.R](./src/Processing/processing_utm10_files.R)

```{r add gemeente and provincie to data_cube}
be_alientaxa_cube <- df %>% 
  left_join(utm1_gemeentes_provincies, by = c("eea_cell_code" = "CELLCODE")) %>% 
  st_drop_geometry() %>% 
  filter(!is.na(isFlanders)|!is.na(isWallonia)|!is.na(isBrussels))

be_alientaxa_cube %>%
  select(c(year,eea_cell_code,taxonKey,n = obs,min_coord_uncertainty,isFlanders,isWallonia,isBrussels,gemeente,provincie,gewest)) %>% 
  write_csv(file.path(cube_path, "be_alientaxa_cube.csv"))
```

Recalculate last_observed column per taxonkey
```{r recalculate_last_observed}
last_observed_taxon <- be_alientaxa_cube %>% 
  group_by(taxonKey) %>% 
  summarise(last_observed_recalc = max(year, na.rm = TRUE),
            isFlanders = any(isFlanders),
            isWallonia = any(isWallonia),
            isBrussels = any(isBrussels),
            isBelgium = TRUE) %>% 
  ungroup() %>% 
  pivot_longer(cols = c(isFlanders, isWallonia, isBrussels, isBelgium), names_to = "region", values_to = "isRegion") %>% 
  filter(isRegion) %>% 
  select(-isRegion) %>% 
  mutate(region = case_match(region,
                             "isFlanders" ~ "Flemish Region",
                             "isWallonia" ~ "Walloon Region",
                             "isBrussels" ~ "Brussels-Capital Region",
                             "isBelgium" ~ "Belgium")) 

data_input_checklist_indicators <- read_tsv("./data/output/UAT_processing/data_input_checklist_indicators.tsv") %>% 
  left_join(last_observed_taxon, by = c("nubKey" = "taxonKey", "locality" = "region")) %>% 
  mutate(last_observed = case_when(is.na(last_observed) ~ last_observed_recalc,
                                   last_observed < last_observed_recalc ~ last_observed_recalc,
                                   TRUE ~ last_observed)) %>% 
  select(-last_observed_recalc) %>% 
  write_tsv("./data/output/UAT_processing/data_input_checklist_indicators.tsv")
```

Free some ram. `be_alientaxa_cube` is no longer needed.
```{r free_ram}
df <- be_alientaxa_cube

rm(be_alientaxa_cube)
gc() #garbage collection
```

The function `createTabularData()` from the alien species portal package is used 
to process & save the data cube to the S3 bucket for use in the alien species portal.

```{r get pre update values}
cube_pre_update <- as.data.frame(aws.s3::get_bucket(bucket = UAT_bucket)) %>% 
  filter(grepl("be_alientaxa_cube_processed.parquet", Key))

cube_pre_update_lastmod <- cube_pre_update$LastModified
cube_pre_update_size <- cube_pre_update$Size
```

```{r save data_cube to S3 bucket}
alienSpecies::createTabularData(dataDir = cube_path, type = "occurrence", bucket = UAT_bucket) 
```

```{r get post update values}
cube_post_update <- as.data.frame(aws.s3::get_bucket(bucket = UAT_bucket)) %>% 
  filter(grepl("be_alientaxa_cube_processed.parquet", Key))

cube_post_update_lastmod <- cube_post_update$LastModified
cube_post_update_size <- cube_post_update$Size

cube_post_update_nrow <- nrow(alienSpecies::loadTabularData(type = "occurrence"))
```

We test if the data have been correctly saved:
```{r test_load_occurrenceData}
test_that("Load occurrenceData", { 
  occurrenceData <- alienSpecies::loadTabularData(type = "occurrence") 
  expect_s3_class( occurrenceData, "data.table") 
}) 

print("lastmod pre update")
print(cube_pre_update_lastmod)
print("lastmod post update")
print(cube_post_update_lastmod)

print("size pre update")
print(cube_pre_update_size)
print("size post update")
print(cube_post_update_size)

print("nrow update")
print(nrow(df %>% filter(year >= 1950 & !is.na(year))))
print("nrow post update")
print(cube_post_update_nrow)

# we expect the post update last modified date to be later than the pre update last modified date
test_that("UPLOAD FAILED: last modified date not updated",
          {
            expect_true(cube_post_update_lastmod > cube_pre_update_lastmod)
          })

# we expect the post update number of rows to be equal to the new number of rows
test_that("UPLOAD FAILED: number of rows not equal",
          {
            expect_equal(cube_post_update_nrow, nrow(df %>% filter(year >= 1950 & !is.na(year))))
          })

df_species <- df %>% 
  filter(year >= 1950 & !is.na(year)) %>% 
  distinct(taxonKey) %>% 
  pull()

post_update_species <- alienSpecies::loadTabularData(type = "occurrence") %>%
  distinct(taxonKey) %>%
  pull()

# we expect all species included in dataset provided to be included in the uploaded data
test_that("UPLOAD FAILED: not all species included",
          {
            
            
            expect_in(df_species, post_update_species)
          })

# we expect 6709291 to be included in both species lists
test_that("UPLOAD FAILED: 6709291 not included",
          {
            expect_true(6709291 %in% df_species)
            expect_true(6709291 %in% post_update_species)
          })

test_that("UPLOAD FAILED: 2502938 not included",
          {
            expect_true(2502938 %in% df_species)
            expect_true(2502938 %in% post_update_species)
          })
```

# Export the cube for timeseries creation

```{r export for artifact upload}
write_csv(df, "./cube_data/be_alientaxa_cube_processed.csv")
```

# Get correct class/kingdom info
```{r get_canonical_names}
taxon_key <-
  df %>%
  distinct(taxonKey) %>%
  pull()
# get basic species information
pb <- progress_bar$new(total = length(taxon_key))
spec_names_raw <- map_df(
  taxon_key,
  function(k) {
    pb$tick()
    name_usage(key = k)$data
  }
) 

spec_names <- spec_names_raw %>%
  mutate(scientificName = coalesce(species, genus, family)) %>% 
  select(
    taxonKey = key,
    canonicalName,
    scientificName,
    rank,
    taxonomicStatus,
    kingdomKey, 
    classKey,
    authorship
  ) %>%
  mutate(canonicalName = ifelse(
    is.na(canonicalName), scientificName, canonicalName
  ))
```

In the rare case multiple taxa share the same scientific name, authorship is added:

```{r add_authorship_if_needed}
spec_names <-
  spec_names %>%
  group_by(canonicalName) %>%
  add_tally() %>%
  ungroup() %>%
  mutate(canonicalName = if_else(n > 1,
                                 scientificName,
                                 canonicalName)) %>%
  select(-c(n))
```

We also retrieve kingdom and class:

```{r get_kingdom_class}
class_key <-
  spec_names %>%
  distinct(classKey) %>%
  filter(!is.na(classKey)) %>%
  pull()
pb <- progress_bar$new(total = length(class_key))
kingdom_class <- map_df(
  class_key,
  function(x) {
    pb$tick()
    name_usage(key = x)$data
  }
) %>%
  select(classKey, class, kingdomKey, kingdom)
```

and we add them:

```{r add_kingdom_class}
# add class
spec_names <-
  spec_names %>%
  left_join(kingdom_class %>%
              distinct(.data$classKey, .data$class),
            by = "classKey"
  )

# add kingdom
spec_names <-
  spec_names %>%
  left_join(kingdom_class %>%
              distinct(.data$kingdomKey, .data$kingdom),
            by = "kingdomKey"
  ) 
```

Preview:

```{r preview_species}
spec_names %>% head(10)
```

Taxa not assigned to any class:

```{r taxa_no_class}
spec_names %>%
  filter(is.na(classKey))
```

export spec_names
```{r}
spec_names %>% 
              distinct(taxonKey, class, classKey, kingdom, kingdomKey) %>% 
  write_csv("./data/spec_names.csv")
```

```{r complete keys}
keys_old <- alienSpecies::loadMetaData(type = "keys")

# backup current keys.csv file
keys_old %>%
  write_csv(here("data", "interim", "keys_old.csv"))

# remove columns we will not use
keys_new <- spec_names %>%
  select(-canonicalName, -authorship)

# rename columns to avoid confusion
colnames(keys_new) <- paste0(colnames(keys_new), "_new")

# Complete knowns, info in cube on checklist
keys_completed <- keys_old %>%
  full_join(keys_new, by = c("scientificName" = "scientificName_new")) %>% 
  mutate(taxonKey = coalesce(taxonKey, taxonKey_new),
         rank = coalesce(rank, rank_new),
         taxonomicStatus = coalesce(taxonomicStatus, taxonomicStatus_new),
         classKey = coalesce(classKey, classKey_new),
         kingdom = coalesce(kingdom, kingdom_new)) %>% 
  select(-contains("new")) %>%
  distinct()

# Find rows with missing taxonKey
keys_missing_taxonkey <- keys_completed %>% 
  filter(is.na(taxonKey))

# Function to derive the most likely taxonKey for a given scientific name
get_derived_taxonKey <- function(scientific_name) {
  taxonomic_data <- tryCatch(
    rgbif::name_usage(name = scientific_name)$data,
    error = function(e) NULL
  )
  if (is.null(taxonomic_data) || nrow(taxonomic_data) == 0) return(NA_integer_)
  
  derived_taxonKey <- taxonomic_data %>% 
    dplyr::filter(taxonomicStatus == "ACCEPTED", !is.na(nubKey)) %>% 
    group_by(nubKey) %>% 
    add_tally() %>% 
    distinct(nubKey, n) %>% 
    ungroup() %>% 
    filter(n == max(n, na.rm = TRUE)) %>% 
    pull(nubKey)
  
  # If multiple, take the first
  if (length(derived_taxonKey) > 0) return(derived_taxonKey[1])
  else return(NA_integer_)
}

# Use lapply to process all missing names
derived_keys <- pbapply::pblapply(keys_missing_taxonkey$scientificName, get_derived_taxonKey)
names(derived_keys) <- keys_missing_taxonkey$scientificName

# Update keys_completed with the derived taxonKeys
keys_completed <- keys_completed %>%
  mutate(
    taxonKey = ifelse(
      is.na(taxonKey) & scientificName %in% names(derived_keys),
      unlist(derived_keys[scientificName]),
      taxonKey
    )
  ) %>% 
  select(taxonKey, scientificName, rank, taxonomicStatus, kingdom, includes)

keys_path <- here("data", "output", "UAT_processing", "be_alientaxa_info.csv")

# Complete unknowns, info not on cube or checklist
## Missing taxonkeys 
# => might not be an issue? => monitoring for now
# Fixing this will be a pain in the ass since `name_usage` returns every use of the scientific name in the gbif backbone as well as on very checklist worldwide
keys_missing_taxonkey <- keys_completed %>% 
  filter(is.na(taxonKey)) %>% 
  write_csv(here("data", "interim", "keys_missing_taxonkey.csv"))

## Missing scientific names => cause of https://github.com/inbo/alien-species-portal/issues/140
keys_missing_name <- keys_completed %>% 
  filter(is.na(scientificName)) %>% 
  write_csv(here("data", "interim", "keys_missing_name.csv"))

# test whether any missing scientificnames remain
test_that("missing scientificNames",
          {
            expect_false(nrow(keys_missing_name) > 0)
          })

keys_completed <- keys_completed %>% 
  mutate(taxonKey = case_when(scientificName == "Trachemys scripta elegans" & is.na(taxonKey) ~ 6157026,
                              scientificName == "Pyrus communis communis" & is.na(taxonKey) ~ 7271631,
                              TRUE ~ taxonKey)) %>% 
  filter(!(scientificName %in% keys_missing_taxonkey$scientificName & is.na(taxonKey)),
         !(taxonKey %in% keys_missing_name$taxonKey & is.na(scientificName))) %>% 
  distinct(taxonKey, scientificName, .keep_all = TRUE)

write_delim(keys_completed, keys_path, ",", quote = "needed")

# Upload to S3 bucket
# aws.s3::put_object(file = keys_path,
#                    object = "keys.csv",
#                    bucket = UAT_bucket,
#                    multipart = TRUE,
#                    show_progress = TRUE,
#                    region = "eu-west-1") 
```